---
title: "Untitled"
date: "`r Sys.Date()`"
output: html_document
---

# connect R to AACT

```{r}
# install.packages("RPostgreSQL")
library(RPostgreSQL)
drv <- dbDriver('PostgreSQL')

con <- dbConnect(drv, dbname="aact",host="aact-db.ctti-clinicaltrials.org", port=5432, user="missionpossible", password="project4")

# example
aact_sample <- dbGetQuery(con, "select distinct study_type from studies")
```

## find most popular disease in the past 10 years excluding COVID-19 and healthy

```{r}
# top 10 most popular conditions for trials start between 2020.01.01 and 2024.12.31
dbGetQuery(con, "select name, count(*) as n from conditions where nct_id in (select nct_id from studies where start_date > '2020-01-01' and start_date < '2024-12-31') group by name order by n desc limit 15")

# extract the name of top 10 condition for trials start between 2020.01.01 and 2024.12.31
top_conditions <- dbGetQuery(con, "select name, count(*) as n from conditions where nct_id in (select nct_id from studies where start_date > '2020-01-01' and start_date < '2024-12-31') group by name order by n desc limit 15")

# count the total number of trials start between 2020.01.01 and 2024.12.31
dbGetQuery(con, "select count(*) from studies where start_date > '2020-01-01' and start_date < '2024-12-31'")


# count the total number of trials where condition is among the top 10 most popular conditions for trials start between 2020.01.01 and 2024.12.31
dbGetQuery(con, "select count(distinct nct_id) from conditions where name in 
                      ('Breast Cancer', 'Covid19', 'Obesity', 'Stroke', 'Depression', 
                       'Anxiety', 'Cancer', 'Pain', 'Heart Failure', 'Parkinson Disease')
           and nct_id in (select nct_id from studies where start_date > '2020-01-01' and start_date < '2024-12-31')")
```

Disease selected: ('Breast Cancer', 'Obesity', 'Stroke', 'Depression', 
                       'Anxiety', 'Cancer', 'Pain', 'Heart Failure', 
                       'Parkinson Disease', 'Prostate Cancer')




## select countries
```{r}
# summary table of the top 15 most popular countries for trials between 2020.01.01 and 2024.12.31 and among the top 10 condition and removed = FALSE 
dbGetQuery(con, "select name, count(*) as n from countries where nct_id in 
                      (select nct_id from conditions where name in 
                      ('Breast Cancer', 'Obesity', 'Stroke', 'Depression', 
                       'Anxiety', 'Cancer', 'Pain', 'Heart Failure', 
                       'Parkinson Disease', 'Prostate Cancer')
           and nct_id in (select nct_id from studies where start_date > '2020-01-01' and start_date < '2024-12-31')) and removed = FALSE group by name order by n desc limit 15")

# extract the name of top 15 countries for trials between 2020.01.01 and 2024.12.31 and among the top 10 condition and removed = FALSE
top_countries <- dbGetQuery(con, "select name, count(*) as n from countries where nct_id in 
                      (select nct_id from conditions where name in 
                      ('Breast Cancer', 'Obesity', 'Stroke', 'Depression', 
                       'Anxiety', 'Cancer', 'Pain', 'Heart Failure', 
                       'Parkinson Disease', 'Prostate Cancer')
           and nct_id in (select nct_id from studies where start_date > '2020-01-01' and start_date < '2024-12-31')) and removed = FALSE group by name order by n desc limit 15")
```

countries selected: ('United States', 'China', 'Turkey', 'France', 'Canada', 
               'United Kingdom', 'Italy', 'Spain', 'Germany', 'Netherlands')


# if we want data for last 10 years 

## find most popular disease in the past 10 years excluding COVID-19 and healthy

```{r}
# top 10 most popular conditions for trials start between 2020.01.01 and 2024.12.31
dbGetQuery(con, "select name, count(*) as n from conditions where nct_id in (select nct_id from studies where start_date > '2015-01-01' and start_date < '2024-12-31') group by name order by n desc limit 15")

# extract the name of top 10 condition for trials start between 2020.01.01 and 2024.12.31
top_conditions <- dbGetQuery(con, "select name, count(*) as n from conditions where nct_id in (select nct_id from studies where start_date > '2015-01-01' and start_date < '2024-12-31') group by name order by n desc limit 15")

# count the total number of trials start between 2020.01.01 and 2024.12.31
dbGetQuery(con, "select count(*) from studies where start_date > '2020-01-01' and start_date < '2024-12-31'")


# count the total number of trials where condition is among the top 10 most popular conditions for trials start between 2020.01.01 and 2024.12.31
dbGetQuery(con, "select count(distinct nct_id) from conditions where name in 
                      ('Breast Cancer', 'Obesity', 'Stroke', 'Depression', 
                       'Pain', 'Anxiety', 'Cancer', 'Heart Failure', 
                       'Prostate Cancer', 'Hypertension')
           and nct_id in (select nct_id from studies where start_date > '2015-01-01' and start_date < '2024-12-31')")
```

Disease selected: ('Breast Cancer', 'Obesity', 'Stroke', 'Depression', 
                       'Pain', 'Anxiety', 'Cancer', 'Heart Failure', 
                       'Prostate Cancer', 'Hypertension')

## select countries
```{r}
# summary table of the top 15 most popular countries for trials between 2015.01.01 and 2024.12.31 and among the top 10 condition and removed = FALSE 
dbGetQuery(con, "select name, count(*) as n from countries where nct_id in 
                      (select nct_id from conditions where name in 
                      ('Breast Cancer', 'Obesity', 'Stroke', 'Depression', 
                       'Pain', 'Anxiety', 'Cancer', 'Heart Failure', 
                       'Prostate Cancer', 'Hypertension')
           and nct_id in (select nct_id from studies where start_date > '2015-01-01' and start_date < '2024-12-31')) and removed = FALSE group by name order by n desc limit 15")

# extract the name of top 15 countries for trials between 2020.01.01 and 2024.12.31 and among the top 10 condition and removed = FALSE
top_countries <- dbGetQuery(con, "select name, count(*) as n from countries where nct_id in 
                      (select nct_id from conditions where name in 
                      ('Breast Cancer', 'Obesity', 'Stroke', 'Depression', 
                       'Pain', 'Anxiety', 'Cancer', 'Heart Failure', 
                       'Prostate Cancer', 'Hypertension')
           and nct_id in (select nct_id from studies where start_date > '2015-01-01' and start_date < '2024-12-31')) and removed = FALSE group by name order by n desc limit 15")

# count the number of trials in the top 10 countries between 2015.01.01 and 2024.12.31 and among the top 10 condition and removed = FALSE
dbGetQuery(con, "select count(distinct nct_id) from countries where name in 
                      ('United States', 'China', 'France', 'Canada', 
                       'United Kingdom', 'Turkey', 'Spain', 'Italy', 
                       'Germany', 'Korea, Republic of')
           and nct_id in (select nct_id from conditions where name in 
                      ('Breast Cancer', 'Obesity', 'Stroke', 'Depression', 
                       'Pain', 'Anxiety', 'Cancer', 'Heart Failure', 
                       'Prostate Cancer', 'Hypertension')
           and nct_id in (select nct_id from studies where start_date > '2015-01-01' and start_date < '2024-12-31')) and removed = FALSE")
```

countries selected ('United States', 'China', 'France', 'Canada', 
               'United Kingdom', 'Turkey', 'Spain', 'Italy', 
               'Germany', 'Korea, Republic of')







# Extract data from database 

## Extract the nct_id

```{r}
# # extract the nct_id of trials between 2020.01.01 and 2024.12.31 and among the top 10 condition, and no missing value in phase, and top 10 countries
# nct_id_top10 <- dbGetQuery(con, "select distinct nct_id from conditions where name in 
#                       ('Breast Cancer', 'Obesity', 'Stroke', 'Depression', 
#                        'Anxiety', 'Cancer', 'Pain', 'Heart Failure', 
#                        'Parkinson Disease', 'Prostate Cancer')
#            and nct_id in (select nct_id from studies where start_date > '2020-01-01' and start_date < '2024-12-31') 
#            and nct_id in (select nct_id from studies where phase is not null) 
#            and nct_id in (select nct_id from countries where name in 
#                       ('United States', 'Canada', 'Germany', 'France', 'United Kingdom', 
#                        'Italy', 'Spain', 'Australia', 'Netherlands', 'China') and removed = FALSE)")

# extract the nct_id of trials between 2015.01.01 and 2024.12.31 and among the top 10 condition, and no missing value in phase, and top 10 countries
nct_id_top10 <- dbGetQuery(con, "select distinct nct_id from conditions where name in 
                      ('Breast Cancer', 'Obesity', 'Stroke', 'Depression', 
                       'Pain', 'Anxiety', 'Cancer', 'Heart Failure', 
                       'Prostate Cancer', 'Hypertension')
           and nct_id in (select nct_id from studies where start_date > '2015-01-01' and start_date < '2024-12-31') 
           and nct_id in (select nct_id from studies where phase is not null) 
           and nct_id in (select nct_id from countries where name in 
                      ('United States', 'China', 'France', 'Canada', 
                       'United Kingdom', 'Turkey', 'Spain', 'Italy', 
                       'Germany', 'Korea, Republic of') and removed = FALSE)")
```

## Query the database for the trials data

```{r}
library(tidyverse)

# Get the nct_id values of interest
nct_id_values <- nct_id_top10$nct_id

# Create placeholders for SQL query
placeholders <- paste(shQuote(nct_id_values, type = "sh"), collapse = ", ")

# Query the studies table
studies_query <- paste0("
  SELECT nct_id, start_date, phase
  FROM studies
  WHERE nct_id IN (", placeholders, ")
")
studies <- dbGetQuery(con, studies_query)

# Query the conditions table
conditions_query <- paste0("
  SELECT name, nct_id
  FROM conditions
  WHERE nct_id IN (", placeholders, ")
")
conditions <- dbGetQuery(con, conditions_query)
# remove all rows in conditions if column name is not in top ten conditions
conditions_filtered <- conditions %>% 
  filter(name %in% c('Breast Cancer', 'Obesity', 'Stroke', 'Depression', 
                       'Pain', 'Anxiety', 'Cancer', 'Heart Failure', 
                       'Prostate Cancer', 'Hypertension'))
# # check if there is one nct_id have multiple conditions - yes 
# conditions_filtered %>% 
#   group_by(nct_id) %>% 
#   summarise(n = n()) %>% 
#   filter(n > 1)

# Query the countries table
countries_query <- paste0("
  SELECT name, nct_id
  FROM countries
  WHERE nct_id IN (", placeholders, ") AND removed = FALSE
")
countries <- dbGetQuery(con, countries_query)
# remove all countries that is not in the top ten countries
countries_filtered <- countries %>% 
  filter(name %in% c('United States', 'China', 'France', 'Canada', 
               'United Kingdom', 'Turkey', 'Spain', 'Italy', 
               'Germany', 'Korea, Republic of'))
```

## data cleaning

### year
```{r}
# create a new column in studies named year, which extract the year information from start_date
studies$year <- as.numeric(format(as.Date(studies$start_date), "%Y"))
```

### condition
```{r}
# Convert the dataset to the desired format
conditions_spread <- conditions_filtered %>%
  mutate(value = 1) %>%  # Create a placeholder column
  pivot_wider(
    names_from = name,   # Make columns from 'name' values
    values_from = value, # Populate columns using the placeholder
    values_fill = 0      # Fill missing combinations with 0
  )
```

### country
```{r}
# Convert the dataset to the desired format
countries_spread <- countries_filtered %>%
  mutate(value = 1) %>%  # Create a placeholder column
  pivot_wider(
    names_from = name,   # Make columns from 'name' values
    values_from = value, # Populate columns using the placeholder
    values_fill = 0      # Fill missing combinations with 0
  )
```

## combining dataset
```{r}
# Combine countries_spread and conditions_spread and studies based on nct_id
data <- studies %>%
  full_join(conditions_spread, by = "nct_id") %>%
  full_join(countries_spread, by = "nct_id")

# write csv file
write.csv(data, "data.csv")
```










